Global spirals in self-gravitating dead zones in protoplanetary discs

### 2014-10-30 paper/model.tex
minor updates
update

Our fiducial parameter values are: $s=2$, $R_{d1}=R_0$, $R_{d2}=2R_0$,
$\epsilon=0.1$, $\delta R_d=5$, $h=0.05$ and
$Q_\mathrm{out}=2$. An example of the initial surface density and the
Toomre parameter (see\ref{wkb}) is shown in Fig. \ref{initial_surf}. 
Initially there is no vertical or radial velocity
($v_R = v_r = v_\theta = 0$). The azimuthal velocity is initialized to
satisfy centrifugal balance with pressure and gravity,
\begin{align}
  \frac{v_\phi^2}{r} = \frac{1}{\rho}\frac{\p p}{\p r} + \frac{\p
    \Phi_\mathrm{tot}}{\p r}
\end{align}
and similarly in 2D. The angular velocity is $\Omega = v_\phi/R$.

@@

The non-axisymmetric surface density can be visualised with
\begin{align}
  \Delta \Sigma = \frac{\Sigma - \Sigma_{0}}{\Sigma_{0}}, 
\end{align}
while an individual mode as
\begin{align}
  \Delta\Sigma_m = \frac{2}{\Sigma_{00}} \real\left[\Sigma_m \exp{(\ii
      m\phi)}\right]
	\end{align}
where $\Sigma_{00} = \Sigma_0(t=0)$.

The global mode amplitude is 
\begin{align}
  A_m = \frac{|C_m|}{M_{d0}}
\end{align}
where $M_{d0}$ is the initial disc mass and 
\begin{align}
  C_m = 2\pi\int_{R_1}^{R_2} \Sigma_m RdR, 
\end{align}
with integration limits chosen later.

The time evolution of the $m^\mathrm{th}$ mode can be characterized by
assuming 
$\Sigma_m\propto\exp{(\ii \sigma t)}$, where
$\sigma = \omega - \ii\gamma$ is the complex frequency, $\omega$ is
the real frequency and $\gamma$ is the growth rate.
We estimate the growth rate as
\begin{align}
  \gamma = \frac{d\ln{A_m}}{dt},
\end{align}
and we estimate the real frequency by calculating the pattern speed
$\Omega_p = \omega/m$,
\begin{align}
  \Omega_p = \frac{d\phi_\mathrm{max}}{dt},
\end{align}
where $\phi_\mathrm{max}$ is the azimuthal co-ordinate of
$\mathrm{max}(\Delta\Sigma_m)$. 
%\begin{align}
%  \mathrm{max}\left\{2 \real\left[ \Sigma_m \exp{(\ii m
%        \phi)}\right]\right\},

>>>>

Our fiducial parameter values are: $s=2$, $R_{d1}=R_0$, $R_{d2}=2R_0$,
$\epsilon=0.1$, $\delta R_d=5$, $h=0.05$ and
$Q_\mathrm{out}=2$. An example of the initial surface density and the
Toomre parameter (see\S\ref{wkb}) is shown in Fig. \ref{initial_surf}. 
Initially there is no vertical or radial velocity
($v_R = v_r = v_\theta = 0$). The azimuthal velocity is initialized to
satisfy centrifugal balance with pressure and gravity,
\begin{align}
  \frac{v_\phi^2}{r} = \frac{1}{\rho}\frac{\p p}{\p r} + \frac{\p
    \Phi_\mathrm{tot}}{\p r}
\end{align}
and similarly in 2D. The angular velocity is $\Omega = v_\phi/R$.

@@

% The non-axisymmetric surface density can be visualised with
% \begin{align}
%   \Delta \Sigma = \frac{\Sigma - \Sigma_{0}}{\Sigma_{0}}, 
% \end{align}
% while an individual mode as
% \begin{align}
%   \Delta\Sigma_m = \frac{2}{\Sigma_{00}} \real\left[\Sigma_m \exp{(\ii
%       m\phi)}\right]
% 	\end{align}
% where $\Sigma_{00} = \Sigma_0(t=0)$.

% The global mode amplitude is 
% \begin{align}
%   A_m = \frac{|C_m|}{M_{d0}}
% \end{align}
% where $M_{d0}$ is the initial disc mass and 
% \begin{align}
%   C_m = 2\pi\int_{R_1}^{R_2} \Sigma_m RdR, 
% \end{align}
% with integration limits chosen later.

The time evolution of the $m^\mathrm{th}$ mode can be characterized by
assuming $\Sigma_m\propto\exp{(\ii \sigma t)}$, where
$\sigma = \omega - \ii\gamma$ is the complex frequency, $\omega$ is
the real frequency and $\gamma$ is the growth rate. In linear theory
$\sigma$ is a constant eigenvalue. 
To estimate $\sigma$ from simulation data, we first construct the real
surface density 
\begin{align}
  \Delta\Sigma_m = \frac{2}{\Sigma_{00}} \real\left[\Sigma_m \exp{(\ii
      m\phi)}\right]
\end{align}
where $\Sigma_{00} = \Sigma_0(t=0)$. We then locate the maximum,
\begin{align}
  \mathrm{max}(\Delta\Sigma_m) \equiv
  \Delta\Sigma_m(R_\mathrm{peak},\phi_\mathrm{peak}). 
\end{align}
We estimate the growth rate as
\begin{align}
  \gamma = \frac{d}{dt}\ln{\left[\mathrm{max}(\Delta\Sigma_m)\right]}, 
\end{align}
and we estimate the real frequency by calculating the pattern speed
$\Omega_p = \omega/m$,
\begin{align}
  \Omega_p = \frac{d\phi_\mathrm{peak}}{dt}. 
\end{align}
The total non-axisymmetric surface density can be visualised with
\begin{align}
  \Delta\Sigma = \frac{\Sigma - \Sigma_0}{\Sigma_0}. 
\end{align}

%\begin{align}
%  \mathrm{max}\left\{2 \real\left[ \Sigma_m \exp{(\ii m
%        \phi)}\right]\right\},

### 2014-10-30 paper/results.tex
update

\section{Preliminary experiments}
Here we describe one numerical simulation from each code to give a
qualitative picture of disc evolution. In the next section, we analyse
2D simulations in more detail and explore parameter space.

In these simulations we subject the disc to random perturbations in
cylindrical velocity,
\begin{align}\label{randpert}
  \frac{\delta v_R}{c_s} = \frac{\delta}{M}\times T(R) \sum_{m=1}^M\cos{m\phi},
\end{align}
where $\delta$ is an amplitude and 
\begin{align}
  T(R) =
  \exp{\left[-\frac{1}{2}\left(\frac{R-\overline{R}_d}{\Delta
          R_d}\right)^2\right]}, 
\end{align}
where $\overline{R}_d = (R_{d1}+R_{d2})/2$ and $\Delta R_d =
(R_{d2}-R_{d1})/2$. 
\subsection{FARGO 2D}\label{fargo_fiducial}
We consider a disc size $[R_\mathrm{min}, R_\mathrm{max}] =
[0.4,10]R_0$. This gives a total disc mass $M_{d}=0.086M_*$, and the
mass within $R\in[R_{d1},R_{d2}]$ is $0.049M_*$. 
We use a resolution of $N_R\times N_\phi = 512\times 1024$ or
about $8$ grids per $H$, and adopt $\epsilon_g=10^{-4}H$ for the 
self-gravity softening length\footnote{In 2D self-gravity, $\epsilon_g$ also
  approximates for the vertical disc thickness, so a more appropriate
  value would be $\epsilon_g\sim H$ \citep{muller12}. However, because
  $\epsilon_g\propto R$ is needed in FARGO, the Poisson kernel
  (Eq. \ref{2d_grav}) is no longer symmetric in $(R,R^\prime)$. We
  choose a small  
  $\epsilon_g$ in favour of angular momentum conservation, keeping in
  mind that the strength of self-gravity will be over-estimated.}.
Perturbations are set initially with $M=10$ and
$\delta\in[-10^{-3},10^{-3}]$ chosen randomly but does not depend
on $\phi$.

%set $M=10$ for random perturbations (Eq. \ref{randpert}).

Snapshots from the disc evolution are shown in Fig. \ref{fargo_2d} in
terms of the non-axisymmetric surface density. At early times
$t\lesssim100P_0$ the disc is dominated by low-amplitude high-$m$
perturbations centred in the dead zone. Eventually a $m=1$ spiral 
grows and dominates inside the dead zone.

Fig. \ref{fargo_2d_angmom} compares the disc angular momenta
evolution, most of which is contained in the $m=0$ and $m=1$ modes. 
The $m=1$ spiral has an associated negative angular
momentum. Its growth is correlated with an increase in the axisymmetric
component of angular momentum, such that $\Delta J_0 \sim - \Delta
J_1$. Note that FARGO does not explicitly conserve angular
momentum. In addition, there may be an angular momentum flux due to
self-gravity across disc boundaries since our domain size is finite. 
Nevertheless, we find the total angular momentum is conserved to 
$\Delta J = O(10^{-7}J)$, and is much smaller in magnitude than the change in the
angular momenta components, $|\Delta J_m|$. 
Fig. \ref{fargo_2d_angmom} then indicates a transfer of 
angular momentum from the $m=1$ spiral to the background disc.

\begin{figure}
  \includegraphics[scale=0.27]{figures/polarxy_dens050}\includegraphics[scale=0.27,clip=true,trim=2.26cm 
  0cm 0cm
  0cm]{figures/polarxy_dens100}\includegraphics[scale=0.27,clip=true,trim=2.26cm
  0cm 0cm 0cm]{figures/polarxy_dens125} 
  \caption{Fiducial FARGO 2D simulation (\S\ref{fargo_fiducial}),
    showing the evolution of the non-axisymmetric surface density
    $\Delta\Sigma$. Note that the disc extends to $R=10R_0$, but only
    the inner part is shown for clarity. \label{fargo_2d}} 
\end{figure}

\begin{figure}
  \includegraphics[width=\linewidth]{figures/nonaxi_evol_ang_fargo}
  \caption{Evolution of angular momentum components in the FARGO
    fiducial simulation (Fig. \ref{fargo_2d}). The perturbation
    relative to $t=0$ is shown in units of the initial total angular
    momentum $J_\mathrm{ref}$.\label{fargo_2d_angmom}} 
\end{figure}

\subsection{ZEUS-MP 3D}
The 3D disc has radial size 
$[r_\mathrm{min},r_\mathrm{max}]=[0.4,10]R_0$ and vertical extent
$n_H=2$ scale-heights. The resolution is $N_r\times N_\theta\times
N_\phi=256\times32\times256$. Because of the much reduced resolution
compared to 2D, we use a smooth perturbation by setting
$\delta = 10^{-3}$ and $M=1$. This corresponds to a single $m=1$
spiral in the dead zone, which was found to dominate the 2D
simulation.

Our 3D discs are initialised in approximate equilibrium only, so we
first evolve the disc without perturbations using  
$(\lmax,\mmax)=(32,0)$ up to $t=10P_0$, during which 
meridional velocities are damped out. We then restart the simulation
with the above perturbation and $(\lmax,\mmax)=(32,32)$.

\subsection{PLUTO 3D}

>>>>

\section{Results for two-dimensional discs}
We first present results from FARGO simulations. The 2D disc spans
$[R_\mathrm{min}, R_\mathrm{max}] = [0.4,10]R_0$. This gives a total
disc mass $M_{d}=0.086M_*$. The mass within
$R\in[R_\mathrm{min},R_{d1}]$ is $0.017M_*$, that within
$R\in[R_{d1},R_{d2}]$ is $0.049M_*$, and that within
$R\in[R_{d2},R_\mathrm{max}]$ is $0.021M_*$. We use a resolution of
$N_R\times N_\phi = 1024\times 2048$, or about $16$ grids per $H$, and
adopt $\epsilon_g=10^{-4}H$ for the   
self-gravity softening length\footnote{In 2D self-gravity, $\epsilon_g$ also
  approximates for the vertical disc thickness, so a more appropriate
  value would be $\epsilon_g\sim H$ \citep{muller12}. However, because
  $\epsilon_g\propto R$ is needed in FARGO, the Poisson kernel
  (Eq. \ref{2d_grav}) is no longer symmetric in $(R,R^\prime)$. We
  choose a small  
  $\epsilon_g$ in favour of angular momentum conservation, keeping in
  mind that the strength of self-gravity will be over-estimated.}.

\subsection{Reference run with random perturbations}
In this fiducial simulation we subject the disc to initial perturbations in
cylindrical radial velocity, 
\begin{align}\label{randpert}
  \frac{\delta v_R}{c_s} = \frac{\delta}{M}\times T(R) \sum_{m=1}^M\cos{m\phi},
\end{align}
where $\delta$ is an amplitude and 
\begin{align}
  T(R) =
  \exp{\left[-\frac{1}{2}\left(\frac{R-\overline{R}_d}{\Delta
          R_d}\right)^2\right]}, 
\end{align}
where $\overline{R}_d = (R_{d1}+R_{d2})/2$ and $\Delta R_d =
(R_{d2}-R_{d1})/2$. We use $M=10$ and choose
$\delta\in[-10^{-3},10^{-3}]$ randomly but independent of $\phi$.

Fig. \ref{fargo_modeamp} plots evolution of the maximum
non-axisymmetric surface density amplitudes in $R\in[R_{d1},R_{d2}]$
for $m\in[1,10]$. Snapshots from the simulation are shown in
Fig. \ref{fargo_2d}. 
At early times $t\lesssim100P_0$ the dead zone is
dominated by low-amplitude high-$m$ perturbations. The $m\geq4$ modes
growth initially and saturate (or decays) after $t=40P_0$. Notice the
low $m\leq 2$ modes decay initially, but grows between $t\in[20,40]P_0$,
possibily due to saturation of the high-$m$ modes  
\citep{laughlin96,laughlin97}. However, the $m=1$ mode begins to grow
again after $t=70P_0$, and eventually dominates the dead zone.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/nonaxi_evol_DZ_fargo}
  \caption{Evolution of maxima in non-axisymmetric surface density 
    in the fiducial FARGO simulation.\label{fargo_modeamp}} 
\end{figure}

\begin{figure}
  \includegraphics[scale=0.27]{figures/polarxy_dens050}\includegraphics[scale=0.27,clip=true,trim=2.26cm
  0cm 0cm
  0cm]{figures/polarxy_dens110}\includegraphics[scale=0.27,clip=true,trim=2.26cm
  0cm 0cm 0cm]{figures/polarxy_dens130} 
  \caption{Visualisation of the FARGO 2D simulation. The total  
    non-axisymmetric surface density
    $\Delta\Sigma$ is shown. \label{fargo_2d}} 
\end{figure}

Fig. \ref{2d_angmom} shows the disc angular momenta
evolution. Only the $m=0,\,1$ components are 
plotted since they are dominant. The $m=1$ structure has
an associated negative angular momentum.   
Its growth is compensated by an increase in the axisymmetric
component of angular momentum, such that $\Delta J_0 + \Delta
J_1 \sim 0$. Note that FARGO does not conserve angular momentum
exactly. However, we find the total angular momentum varies by 
$|\Delta J/J|= O(10^{-6})$, and is much smaller in magnitude than the
change in the angular momenta components, $|\Delta J_{0,1}/J|>
O(10^{-5})$. Fig. \ref{2d_angmom} then suggest that angular momentum
is transferred from the one-arm spiral to the background disc.

\begin{figure}
  % scale=0.41
  \includegraphics[scale=.41]{figures/nonaxi_evol_ang_fargo}
  \caption{Evolution of angular momentum components in the fiducial
    FARGO simulation. The perturbation
    relative to $t=0$ in 2D is shown in units of the
    initial total angular momentum $J_\mathrm{ref}$.\label{2d_angmom}} 
\end{figure}

\subsection{Non-linear initial perturbation}
In order to focus on the one-arm spiral that emerges in the above
simulation, we here analyse a run with only an initial $m=1$
perturbation in the dead zone. Specifically, we set
\begin{align}
  &\Sigma \to \Sigma_0(R) + \delta\Sigma_0(R) + 2\Sigma_1(R)\cos{\phi},\\
  &v_\phi  \to v_{\phi 0}(R) + 2v_{\phi 1}(R)\cos{\phi},
\end{align}
for $R\in[R_{d1},R_{d2}]$, where
\begin{align}
  &\delta\Sigma_0 =
  -\delta^2\times\Sigma_0(R_{d1})\frac{R_{d1}}{R}\sin^3{\left[\frac{2\pi\left(R-R_{d1}\right)}{R_{d2}-R_{d1}}\right]},\\ 
  &\Sigma_1 =
  \delta\times\Sigma_0\sin{\left[\frac{\pi\left(R-R_{d1}\right)}{R_{d2}-R_{d1}}\right]},  
\end{align}
with $\delta = 10^{-3}$, and
\begin{align}
  v_{\phi 1} = -\frac{v_{\phi 0}\delta\Sigma_0}{2\Sigma_1}. 
\end{align}
This perturbs the dead zone structure while conserving total mass and
angular momentum. The $m=1$ perturbation has an associated negative
angular momentum (as observed in the previous simulation) at the
expense of increasing the angular momentum associated with the
background disc. %local exchange for simplicity 
%actually doesn't matter if we set plus or minus
Note that the perturbation to the axisymmetric surface density is
$O(\delta^2)$, whereas that in $m=1$ is $O(\delta)$. Since $Q>1$
everywhere in the disc, there is no risk of axisymmetric gravitational
instability.

While the above form of initial perturbations is artificial, we find 
the spiral that forms have similar properties (see below) to that found in the
previous simulation. However, the above perturbations have the
advantage of producing cleaner results without complications from
higher $m$ modes, and the required simulation time for the one-arm
spiral to emerge is significantly reduced.

Fig. shows the evolution of angular momenta.
